% D-declaration-flow.tex — Annex D: Declaration Flow
\chapter{Declaration Flow}
\label{annex:D-declaration-flow}
\index{declaration flow}
\index{declarations!lifecycle}

\pnum
This annex traces the lifecycle of a Lain declaration from source text to
emitted C, following a single \lain{func} declaration as the running example.
The same pipeline applies to \lain{proc}, \lain{struct}, \lain{enum},
\lain{import}, \lain{extern func}, \lain{extern type}, and \lain{type} aliases,
with per-kind differences noted where they diverge.

% -------------------------------------------------------------------------
\section{Stage 1: Lexing}
\label{sec:df-lex}
% -------------------------------------------------------------------------

\pnum
The lexer (\srcref{src/lexer.h}) emits a stream of \cname{Token} values.
For a declaration, the relevant tokens are:
\begin{itemize}
\item \cname{TOKEN\_KEYWORD\_FUNC} or \cname{TOKEN\_KEYWORD\_PROC} — the
  declaration introducer.
\item \cname{TOKEN\_IDENTIFIER} — the function name.
\item \cname{TOKEN\_LPAREN} / \cname{TOKEN\_RPAREN} — parameter list delimiters.
\item Type-name identifiers and type-qualifier keywords
  (\cname{TOKEN\_KEYWORD\_MUT}, \cname{TOKEN\_KEYWORD\_MOV}) for parameters
  and the return type.
\item \cname{TOKEN\_LBRACE} / \cname{TOKEN\_RBRACE} — body delimiters.
\end{itemize}

\pnum
The lexer does not attach meaning to any token.  The stream is consumed
on-demand by the parser via \cname{lexer\_next()} calls.

% -------------------------------------------------------------------------
\section{Stage 2: Parsing}
\label{sec:df-parse}
% -------------------------------------------------------------------------

\pnum
\cname{parse\_decl} (\srcref{src/parser/decl.h}) dispatches on the leading
keyword.  For \cname{TOKEN\_KEYWORD\_FUNC}, it calls
\cname{parse\_func\_proc\_decl\_impl} with \cname{is\_proc = false}.

\pnum
The parser allocates a \cname{Decl} node on \cname{ast\_arena} via
\cname{ARENA\_PUSH\_STRUCT}.  It fills in:
\begin{itemize}
\item \cname{kind = DECL\_FUNCTION} (or \cname{DECL\_PROCEDURE}).
\item \cname{name} — the identifier string (pointer into source, zero-copy).
\item \cname{as.function\_decl.params} — a \cname{DeclList *} of
  \cname{DECL\_VARIABLE} nodes, one per parameter.
\item \cname{as.function\_decl.return\_type} — a \cname{Type *} constructed
  from the declared return type tokens.
\item \cname{as.function\_decl.body} — a \cname{StmtList *} of body statements,
  recursively produced by \cname{parse\_stmt\_list}.
\item \cname{line}, \cname{col} — source position of the declaration keyword.
\end{itemize}

\pnum
For \lain{extern func}, \cname{parse\_extern\_func\_proc\_decl\_impl} is called.
It produces a \cname{DECL\_EXTERN\_FUNCTION} node with \cname{body = NULL}.
For \lain{struct}, \cname{parse\_struct\_decl} fills \cname{as.struct\_decl.fields}.
For \lain{import}, \cname{parse\_import\_decl} fills \cname{as.import\_decl.path}.

% -------------------------------------------------------------------------
\section{Stage 3: Module loading}
\label{sec:df-module}
% -------------------------------------------------------------------------

\pnum
For \cname{DECL\_IMPORT} nodes, \cname{sema\_resolve\_module}
(\srcref{src/sema.h}) calls \cname{load\_module} (\srcref{src/module.h}).
\cname{load\_module}:
\begin{enumerate}
\item Translates the dotted import path to a file path.
\item Reads and lexes the file.
\item Recursively calls the parser to produce a \cname{DeclList *}.
\item Splices the loaded declarations in-place into the caller's declaration
  list, replacing the \cname{DECL\_IMPORT} node.
\item Marks the module path in the loaded-modules list to prevent cycles.
\end{enumerate}

\pnum
After this stage, the \cname{DeclList *} seen by the rest of sema contains only
flattened, non-import declarations.

% -------------------------------------------------------------------------
\section{Stage 4: Resolve phase}
\label{sec:df-resolve}
% -------------------------------------------------------------------------

\pnum
\cname{sema\_resolve\_module} iterates the declaration list twice:

\paragraph{Forward-declaration scan.}
On the first pass, every top-level \cname{DECL\_FUNCTION},
\cname{DECL\_PROCEDURE}, \cname{DECL\_STRUCT}, \cname{DECL\_ENUM},
\cname{DECL\_EXTERN\_FUNCTION}, \cname{DECL\_EXTERN\_PROCEDURE},
\cname{DECL\_EXTERN\_TYPE}, and \cname{DECL\_VARIABLE} is inserted into
\cname{sema\_globals} (\srcref{src/sema/resolve.h} lines 128–200):
\begin{itemize}
\item Functions: a \cname{Symbol} with \cname{kind = SYM\_FUNCTION} and a
  pointer back to the \cname{Decl} node.
\item Struct/enum types: a \cname{Symbol} with \cname{kind = SYM\_TYPE}.
\item Global variables: a \cname{Symbol} with \cname{kind = SYM\_VARIABLE}.
\end{itemize}

\noindent
This forward scan enables mutual calls between functions without requiring
forward declarations in user code.

\paragraph{Body resolution.}
On the second pass, \cname{sema\_resolve\_decl} visits each declaration.
For function bodies, \cname{sema\_resolve\_stmt} is called on every body
statement.  During body resolution:
\begin{itemize}
\item Parameter \cname{DECL\_VARIABLE} nodes are inserted into
  \cname{sema\_locals}.
\item Identifiers in expressions are resolved by \cname{sema\_lookup} (locals
  first, then globals), and the \cname{->decl} pointer on \cname{Expr} nodes is
  set.
\item UFCS desugaring rewrites \cname{EXPR\_MEMBER} calls to prepend the
  receiver as the first argument.
\item Generic calls are expanded: comptime arguments are evaluated, the body
  is deep-cloned, type parameters are substituted, and the specialisation is
  recursively resolved and inserted into \cname{sema\_globals}.
\item \cname{comptime if} branches are evaluated and the non-taken branch is
  pruned.
\end{itemize}

% -------------------------------------------------------------------------
\section{Stage 5: Walk phase}
\label{sec:df-walk}
% -------------------------------------------------------------------------

\pnum
After body resolution, \cname{sema\_walk\_phase} is set to true and
\cname{walk\_stmt} is called on each body statement.  In this phase:

\begin{itemize}
\item \cname{sema\_infer\_expr} propagates types bottom-up, annotating every
  \cname{Expr} node's \cname{->type} field with the inferred \cname{Type *}.
  Integer widening is applied at binary operations.
\item VRA propagates integer ranges for every identifier and expression.
\item In-guard proofs are established at \lain{while} / \lain{if} conditions.
\item Bounds checks are performed at every \cname{EXPR\_INDEX} node.
\item The linearity checker tracks consumption state for linear variables,
  enforcing E001–E008 and E003 at scope exit.
\item The borrow checker enforces aliasing rules (E004) and lifetime
  constraints (E010).
\item The bounded-while verifier checks termination measures (E080–E082).
\end{itemize}

% -------------------------------------------------------------------------
\section{Stage 6: Mutual recursion check}
\label{sec:df-mrec}
% -------------------------------------------------------------------------

\pnum
After all declarations have been resolved and walked, the mutual recursion
DFS (\srcref{src/sema.h}) scans the resolved call graph for cycles.  Any
cycle involving a \lain{func} declaration triggers \diagcode{E011}.

% -------------------------------------------------------------------------
\section{Stage 7: Code emission}
\label{sec:df-emit}
% -------------------------------------------------------------------------

\pnum
The emitter (\srcref{src/emit/core.h}) iterates the fully-resolved and
walk-annotated \cname{DeclList *} in six sub-passes:

\begin{enumerate}
\item \textbf{Runtime header} — \srcref{src/emit/lain\_header.h} is written
  verbatim.
\item \textbf{\texttt{c\_include} directives} — \cname{DECL\_C\_INCLUDE} nodes
  are emitted as \texttt{\#include} lines.
\item \textbf{Type forward declarations} — \srcref{src/emit/type\_order.h}
  topologically sorts struct and enum types and emits forward
  \texttt{typedef struct} and \texttt{typedef enum} declarations to satisfy
  C's requirement that types be declared before use.
\item \textbf{Struct/enum definitions} — full \texttt{typedef struct \{...\}}
  and \texttt{typedef enum \{...\}} bodies are emitted.  For enums with payload
  variants, a tag enum and a tagged-union struct are emitted.  Constructor
  functions are emitted by \srcref{src/emit/ctor.h}.
\item \textbf{Function prototypes} — \cname{emit\_forward\_decl}
  (\srcref{src/emit/decl.h}) emits a C prototype for every
  \cname{DECL\_FUNCTION}, \cname{DECL\_PROCEDURE},
  \cname{DECL\_EXTERN\_FUNCTION}, and \cname{DECL\_EXTERN\_PROCEDURE}.
  This allows the function bodies to reference each other freely.
\item \textbf{Function bodies} — \cname{emit\_decl} emits the full C function
  definition for every \cname{DECL\_FUNCTION} and \cname{DECL\_PROCEDURE}.
  Extern declarations are emitted as prototypes only (no body).
\end{enumerate}

% -------------------------------------------------------------------------
\section{Per-kind summary}
\label{sec:df-summary}
% -------------------------------------------------------------------------

\begin{longtable}{p{3.5cm}p{3.5cm}p{6.8cm}}
\caption{Declaration lifecycle by kind.\label{tab:decl-lifecycle}} \\
\toprule
\textbf{DeclKind} & \textbf{Resolve action} & \textbf{Emit action} \\
\midrule
\endfirsthead
\multicolumn{3}{l}{\small (continued)} \\
\toprule
\textbf{DeclKind} & \textbf{Resolve action} & \textbf{Emit action} \\
\midrule
\endhead
\bottomrule
\endfoot

\cname{DECL\_FUNCTION} /
\cname{DECL\_PROCEDURE} &
Symbol inserted into \cname{sema\_globals}; body resolved and walked;
mutual recursion checked. &
Type forward decl (if needed), prototype, full body via
\cname{emit\_decl}. \\

\cname{DECL\_EXTERN\_FUNCTION} /
\cname{DECL\_EXTERN\_PROCEDURE} &
Symbol inserted into \cname{sema\_globals}; no body to resolve. &
C \texttt{extern} prototype only. \\

\cname{DECL\_STRUCT} &
Symbol inserted into \cname{sema\_globals}; field types resolved. &
Forward typedef, then full \texttt{typedef struct \{...\}}. \\

\cname{DECL\_ENUM} &
Symbol inserted into \cname{sema\_globals}; variant types resolved. &
Forward typedef, tag enum, tagged-union struct, constructor functions. \\

\cname{DECL\_IMPORT} /
\cname{DECL\_EVAL\_IMPORT} &
Replaced in-place by the loaded module's \cname{DeclList *} before resolve. &
Not emitted directly; replaced declarations are emitted. \\

\cname{DECL\_C\_INCLUDE} &
No sema action. &
Emitted as \texttt{\#include "..."} in the includes section. \\

\cname{DECL\_EXTERN\_TYPE} &
Symbol inserted into \cname{sema\_globals} as \cname{SYM\_TYPE}. &
Emitted as a forward C \texttt{typedef} or opaque struct. \\

\cname{DECL\_TYPE\_ALIAS} &
Comptime RHS evaluated; result stored in \cname{DeclTypeAlias.resolved\_type}. &
Emitted as a C \texttt{typedef}. \\

\cname{DECL\_VARIABLE} &
If top-level: inserted into \cname{sema\_globals}.
If parameter: inserted into \cname{sema\_locals} during body resolve. &
Top-level variables emitted as C global declarations;
parameters emitted as part of their function signature. \\

\cname{DECL\_DESTRUCT} &
Expanded during resolve into a hidden \texttt{\_param\_N} parameter plus
assignment statements for each field. &
Not directly emitted; expansion is emitted as assignments. \\

\end{longtable}
