% 07-types.tex — Clause 7: Types
\chapter{Types}
\label{sec:07-types}
\index{type}

\pnum
Lain is statically typed.  Every variable, expression, and parameter has a
type determined at compile time.  There is no runtime type information
(RTTI).  The grammar productions for types are in Annex~A (\gramref{type}).

\section{Type taxonomy}
\label{sec:types.taxonomy}
\index{type!taxonomy}

\pnum
The types of Lain are organized as follows:

\begin{itemize}
  \item \textit{Primitive types}: integer, floating-point, boolean, void
        (\S\,\ref{sec:types.primitive});
  \item \textit{Array types}: fixed-size, compile-time-sized
        (\S\,\ref{sec:types.array});
  \item \textit{Slice types}: dynamically-sized views into contiguous memory
        (\S\,\ref{sec:types.slice});
  \item \textit{Struct types}: named product types (\S\,\ref{sec:types.struct});
  \item \textit{Enum types}: simple enumerations (\S\,\ref{sec:types.enum});
  \item \textit{ADT types}: sum types with associated data (\S\,\ref{sec:types.adt});
  \item \textit{Pointer types}: raw memory addresses, unsafe only
        (\S\,\ref{sec:types.pointer});
  \item \textit{Constrained types}: types with value range annotations
        (\S\,\ref{sec:types.constrained}).
\end{itemize}

\pnum
Types are classified as \textit{linear} or \textit{unrestricted}.  Linear
types must be consumed exactly once (see \S\,\ref{sec:11-ownership}).
Unrestricted types may be used any number of times.

\pnum
All primitive types are unrestricted.  A struct type is linear if and only
if at least one of its fields is of a linear type.  This property is
transitive.

\section{Primitive types}
\label{sec:types.primitive}
\index{primitive type}

\subsection{Integer types}
\label{sec:types.integer}
\index{integer type}

\pnum
The integer types are listed below.  Sizes marked \textit{(platform)} are
\idbehav{determined by the target platform as in the C99 standard}.

\begin{longtable}{@{}lllll@{}}
\toprule
\textbf{Type} & \textbf{Signedness} & \textbf{Bit width} & \textbf{Range} & \textbf{C99 type} \\
\midrule
\endfirsthead
\multicolumn{5}{l}{\textit{(continued)}}\\
\toprule
\textbf{Type} & \textbf{Signedness} & \textbf{Bit width} & \textbf{Range} & \textbf{C99 type} \\
\midrule
\endhead
\bottomrule
\endlastfoot
\lain{u8}    & unsigned & 8    & $[0,\, 2^8-1]$       & \texttt{uint8\_t}  \\
\lain{u16}   & unsigned & 16   & $[0,\, 2^{16}-1]$    & \texttt{uint16\_t} \\
\lain{u32}   & unsigned & 32   & $[0,\, 2^{32}-1]$    & \texttt{uint32\_t} \\
\lain{u64}   & unsigned & 64   & $[0,\, 2^{64}-1]$    & \texttt{uint64\_t} \\
\lain{usize} & unsigned & (platform) & $[0,\, 2^N-1]$ & \texttt{size\_t}   \\
\lain{i8}    & signed   & 8    & $[-2^7,\, 2^7-1]$    & \texttt{int8\_t}   \\
\lain{i16}   & signed   & 16   & $[-2^{15},\, 2^{15}-1]$ & \texttt{int16\_t} \\
\lain{i32}   & signed   & 32   & $[-2^{31},\, 2^{31}-1]$ & \texttt{int32\_t} \\
\lain{i64}   & signed   & 64   & $[-2^{63},\, 2^{63}-1]$ & \texttt{int64\_t} \\
\lain{isize} & signed   & (platform) & $[-2^{N-1},\, 2^{N-1}-1]$ & \texttt{ptrdiff\_t} \\
\lain{int}   & signed   & $\geq32$ & (platform) & \texttt{int} \\
\end{longtable}

\pnum
\lain{int} is the default integer type; it corresponds to C's \texttt{int}
and is at least 32 bits wide.  Its exact width is \idbehav{the width of
C's \texttt{int} on the target platform}.

\pnum
\lain{usize} and \lain{isize} are the platform-native unsigned and signed
integer types whose width equals the pointer width.  Their exact width is
\idbehav{the pointer width on the target platform}.

\subsection{Integer rank and widening}
\label{sec:types.rank}
\index{integer rank}
\index{integer widening}

\pnum
The integer types are partially ordered by \textit{rank}:

\begin{longtable}{@{}lll@{}}
\toprule
\textbf{Rank} & \textbf{Types} & \textbf{Bit width} \\
\midrule
\endfirsthead
\toprule
\textbf{Rank} & \textbf{Types} & \textbf{Bit width} \\
\midrule
\endhead
\bottomrule
\endlastfoot
1 & \lain{u8}, \lain{i8}             & 8 \\
2 & \lain{u16}, \lain{i16}           & 16 \\
3 & \lain{u32}, \lain{i32}, \lain{int} & 32 \\
4 & \lain{u64}, \lain{i64}           & 64 \\
5 & \lain{usize}, \lain{isize}       & platform \\
\end{longtable}

\pnum
\textbf{Implicit widening}: a value of a lower-rank integer type may be
used where a higher-rank integer type is expected, without an explicit cast.
The result type of a binary arithmetic or bitwise operation is the
higher-rank operand type.

\begin{example}
\begin{laincode}
var a u8  = 42
var b int = a          // u8 -> int: implicit widening, OK
var c = a + 1000       // u8 + int -> int (wider wins)
\end{laincode}
\end{example}

\begin{constraint}
  An implicit conversion from a higher-rank integer type to a lower-rank
  integer type (narrowing) is \illformed{}.  Narrowing requires an explicit
  \lain{as} cast (\S\,\ref{sec:08-expressions}).
\end{constraint}

\begin{constraint}
  An implicit conversion between integer types and floating-point types is
  \illformed{}.  Such conversions require an explicit \lain{as} cast.
\end{constraint}

\begin{constraint}
  An implicit conversion between integer types and \lain{bool} is
  \illformed{}.  Such conversions require an explicit \lain{as} cast.
\end{constraint}

\subsection{Floating-point types}
\label{sec:types.float}
\index{floating-point type}

\pnum
\lain{f32} and \lain{f64} are 32-bit and 64-bit floating-point types
conforming to IEC~60559 (IEEE~754).  They correspond to C's \texttt{float}
and \texttt{double} respectively.

\pnum
The default type of a floating-point literal is \lain{f64}
(\S\,\ref{sec:lex.literals.float}).

\pnum
Arithmetic on \lain{f32} and \lain{f64} follows IEEE~754 semantics:
overflow produces $\pm\infty$; the result of $0.0/0.0$ and similar
operations is NaN.  These are defined behaviors in Lain.

\subsection{Boolean type}
\label{sec:types.bool}
\index{boolean type}

\pnum
\lain{bool} has exactly two values: \lain{true} and \lain{false}.
\lain{bool} is stored as a single byte; its exact representation is
\idbehav{whether \lain{true} is 1 or any non-zero value}.

\begin{constraint}
  There are no implicit conversions between \lain{bool} and any integer
  type.  \lain{if 1 \{ \}} is \illformed{}; use \lain{if 1 != 0 \{ \}}.
\end{constraint}

\subsection{The void type}
\label{sec:types.void}
\index{void type}

\pnum
\lain{void} represents the absence of a value.  It is used exclusively
in pointer types (\lain{*void}) for C interoperability (see
\S\,\ref{sec:17-interop}).

\begin{constraint}
  A variable declaration with type \lain{void} is \illformed{}.
\end{constraint}

\section{Array types}
\label{sec:types.array}
\index{array type}

\pnum
An array type \lain{[T; N]} denotes a contiguous sequence of $N$ values of
type~\lain{T}.  $N$ shall be a compile-time constant integer greater than
zero (grammar: \gramref{array-type}).

\pnum
The \texttt{.len} property of an array type yields $N$ as a compile-time
constant of type \lain{int}.

\pnum
Array elements are laid out contiguously in memory in increasing index
order, with element~0 at the base address.  The stride between elements is
\idbehav{equal to \texttt{sizeof(T)} in C99, including any intra-element
padding}.

\begin{constraint}
  An array size $N \leq 0$ is \illformed{}~(\diagcode{E100}).
\end{constraint}

\begin{constraint}
  Every array index expression shall satisfy
  $0 \leq \textit{idx} < \textit{arr.len}$.
  An access for which value range analysis cannot prove this is a
  constraint violation~(\diagcode{E014}).
\end{constraint}

\begin{example}
\begin{laincode}
var arr = [int; 5]{1, 2, 3, 4, 5}
var x = arr[2]         // OK: 2 is statically in [0, 4]
// var y = arr[5]      // [E014]: 5 >= arr.len
\end{laincode}
\end{example}

\section{Slice types}
\label{sec:types.slice}
\index{slice type}

\pnum
A slice type \lain{[T]} is a fat pointer to a contiguous sequence of values
of type~\lain{T} whose length is determined at runtime.  Its runtime
representation is a pair \texttt{(ptr: *T, len: usize)}.

\pnum
\lain{str} is a predefined alias for \lain{[u8]}.

\pnum
A slice has two properties accessible via dot notation:
\begin{itemize}
  \item \texttt{.len}: the number of elements, type \lain{usize};
  \item \texttt{.data}: a pointer to the first element, type \lain{*T}.
\end{itemize}

\pnum
A string literal of $N$ characters has static type \lain{Fixed\_u8\_{N+1}}
(a null-terminated fixed-size array of $N{+}1$ bytes) and is implicitly
coercible to \lain{[u8]}.

\section{Struct types}
\label{sec:types.struct}
\index{struct type}

\pnum
A struct type groups named fields into a single named product type.  The
grammar production is \gramref{struct-declaration} (Annex~A).

\pnum
Field order is the declaration order.  The memory layout of a struct is
\idbehav{equal to the C99 struct layout, including compiler-inserted padding
for alignment}.

\pnum
A struct type~$S$ is linear if at least one field of $S$ has a linear type
(transitively).

\pnum
\textbf{Struct construction}: all fields shall be provided in declaration
order.  The type of each argument shall be compatible with the declared
field type (\S\,\ref{sec:types.compat}).

\begin{constraint}
  A struct constructor expression with a number of arguments different from
  the number of fields is \illformed{}~(\diagcode{E100}).
\end{constraint}

\begin{constraint}
  Writing to a field of a struct variable requires the variable to be
  declared as \lain{var} (mutable).  Writing to an immutable binding's
  field is \illformed{}.
\end{constraint}

\begin{constraint}
  Using \lain{==} or \lain{!=} on a struct type is \illformed{}.
  Field-by-field comparison must be performed explicitly.
\end{constraint}

\section{Enum types}
\label{sec:types.enum}
\index{enum type}

\pnum
An enum type is a type with a finite, named set of values.  Each variant
carries no associated data.  The grammar production is
\gramref{enum-declaration} (Annex~A).

\pnum
The underlying representation of an enum type is \idbehav{an integer type
large enough to hold all variant tags; the reference implementation uses
\texttt{int}}.

\pnum
Enum values are accessed as \lain{TypeName.VariantName}.

\begin{constraint}
  Using \lain{==} or \lain{!=} to compare enum values is \illformed{}.
  Enum values shall be compared using \lain{case} pattern matching
  (\S\,\ref{sec:15-match}).
\end{constraint}

\begin{note}
  The constraint on \lain{==} for enums reflects that the underlying tag
  representation is implementation-defined.  Pattern matching is always
  sound.
\end{note}

\section{Algebraic data types (ADTs)}
\label{sec:types.adt}
\index{algebraic data type}

\pnum
An ADT is a sum type in which each variant may carry associated fields.
A type declaration is classified as an ADT if at least one variant has
fields.  The grammar production is \gramref{enum-declaration} (Annex~A,
using the record-variant form).

\pnum
ADTs are represented in C as a tagged union: a tag field of implementation-
defined integer type, plus a union of variant payload structs.

\pnum
ADT values are constructed as \lain{TypeName.VariantName(args)} for
variants with fields, or \lain{TypeName.VariantName} for unit variants.

\pnum
ADT values are destructured via \lain{case} pattern matching
(\S\,\ref{sec:15-match}).

\section{Pointer types}
\label{sec:types.pointer}
\index{pointer type}

\pnum
A pointer type \lain{*T} or \lain{const *T} is a raw memory address.
Pointer operations require an \lain{unsafe} block (\S\,\ref{sec:18-unsafe}).

\pnum
The ownership modes for pointer types are:

\begin{longtable}{@{}llll@{}}
\toprule
\textbf{Lain syntax} & \textbf{Mode} & \textbf{C99 type} & \textbf{Linear?} \\
\midrule
\endfirsthead
\toprule
\textbf{Lain syntax} & \textbf{Mode} & \textbf{C99 type} & \textbf{Linear?} \\
\midrule
\endhead
\bottomrule
\endlastfoot
\lain{*T}     & shared (read-only) & \texttt{const T*}  & no \\
\lain{var *T} & mutable            & \texttt{T*}        & no \\
\lain{mov *T} & owned              & \texttt{T*}        & yes \\
\end{longtable}

\begin{constraint}
  Dereferencing a pointer or taking the address of a variable shall occur
  only inside an \lain{unsafe} block.  Pointer operations in safe code are
  \illformed{}~(\diagcode{E012}).
\end{constraint}

\section{Constrained types}
\label{sec:types.constrained}
\index{constrained type}

\pnum
A constrained type is a type annotated with a value range constraint:

\begin{center}
\textit{T} \textit{op} \textit{bound}
\end{center}

where \textit{op} $\in \{$\lain{>=}, \lain{<=}, \lain{>}, \lain{<},
\lain{==}$\}$ and \textit{bound} is an integer literal.

\pnum
Constrained types are used in function return types and parameter types to
declare value range invariants that the value range analysis
(\S\,\ref{sec:13-vra}) shall enforce.

\begin{example}
\begin{laincode}
func nonzero_divisor() int >= 1 {
    return 42   // VRA verifies: 42 >= 1 holds
}
\end{laincode}
\end{example}

\pnum
The constraint is part of the function signature, not the type system's
nominal identity.  Two constrained types with the same base type are
assignment-compatible.

\section{Type compatibility and equivalence}
\label{sec:types.compat}
\index{type equivalence}
\index{type compatibility}

\pnum
Lain uses \textbf{nominal type equivalence}: two types are the same if and
only if they have the same declared name.

\pnum
Two types $T_1$ and $T_2$ are \textit{compatible} for an expression context
if:
\begin{itemize}
  \item $T_1 = T_2$ (same nominal type), or
  \item $T_1$ is an integer type of lower rank than $T_2$ (implicit
        widening; \S\,\ref{sec:types.rank}).
\end{itemize}

\begin{constraint}
  An expression of type $T_1$ used in a context requiring type $T_2$, where
  $T_1$ and $T_2$ are not compatible, is \illformed{}.
\end{constraint}

\pnum
The operators \lain{==} and \lain{!=} are defined for:
\begin{itemize}
  \item primitive numeric types;
  \item \lain{bool};
  \item pointer types.
\end{itemize}
They are not defined for struct types, array types, or enum types.
