% 10-declarations.tex — Clause 10: Declarations
\chapter{Declarations}
\label{sec:10-declarations}
\index{declaration}

\pnum
This clause specifies top-level declarations.  Local variable declarations
are specified in \S\,\ref{sec:stmt.var}.  The grammar productions are in
Annex~A (\gramref{function-declaration}, \gramref{struct-declaration},
\gramref{enum-declaration}).

\section{Struct declarations}
\label{sec:decl.struct}
\index{struct declaration}

\pnum
A struct declaration introduces a named product type with one or more
named fields.  Fields are listed in declaration order; order is significant
for construction and layout.

\begin{laincode}
struct Point {
    x int
    y int
}
\end{laincode}

\pnum
A field may be annotated with \lain{mov}, making the field (and therefore
the containing struct) a linear type:

\begin{laincode}
struct File {
    mov handle *FILE
}
\end{laincode}

\begin{constraint}
  A struct with zero fields is \illformed{}~(\diagcode{E100}).
\end{constraint}

\section{Enum and ADT declarations}
\label{sec:decl.enum}
\index{enum declaration}
\index{ADT declaration}

\pnum
An enum declaration introduces a type whose values are drawn from a fixed,
named set of variants.  If every variant has no associated fields, the type
is a simple enum (\S\,\ref{sec:types.enum}).  If at least one variant has
associated fields, the type is an algebraic data type (ADT)
(\S\,\ref{sec:types.adt}).

\begin{laincode}
enum Direction { North, South, East, West }  // simple enum

enum Shape {                                  // ADT
    Circle     { radius int }
    Rectangle  { width int, height int }
    Point
}
\end{laincode}

\section{Function declarations}
\label{sec:decl.func}
\index{function declaration}

\pnum
A function declaration introduces a \lain{func}: a pure, terminating,
side-effect-free callable.  Its grammar production is
\gramref{function-declaration}.

\pnum
The signature consists of: the keyword \lain{func}, a name, an optional
list of \lain{comptime} type parameters, a parameter list, and a return
type annotation.

\subsection{Parameters}
\label{sec:decl.func.params}
\index{parameter}

\pnum
Each parameter has a name, a type, and an ownership mode.  The ownership
mode determines how the argument is passed:

\begin{longtable}{@{}lll@{}}
\toprule
\textbf{Syntax} & \textbf{Mode} & \textbf{ABI (struct T)} \\
\midrule
\endfirsthead
\toprule
\textbf{Syntax} & \textbf{Mode} & \textbf{ABI (struct T)} \\
\midrule
\endhead
\bottomrule
\endlastfoot
\lain{name T}        & shared (read-only) & \texttt{const T*} \\
\lain{var name T}    & mutable            & \texttt{T*}       \\
\lain{mov name T}    & owned (transfer)   & \texttt{T} (by value) \\
\end{longtable}

\begin{note}
  For primitive types (integer, float, bool), the ABI passes the value
  directly regardless of ownership mode, as for primitive C arguments.
\end{note}

\pnum
A parameter declared \lain{comptime name type} is a compile-time type
parameter; it must be bound to a concrete type at every call site.

\subsection{Return type annotation}
\label{sec:decl.func.return}
\index{return type}

\pnum
The return type is written after the parameter list.  It may carry a value
range constraint:

\begin{laincode}
func abs(x int) int >= 0 { ... }
\end{laincode}

\begin{constraint}
  All return paths of a \lain{func} shall produce a value whose type and
  value range satisfy the declared return type and constraint.
\end{constraint}

\subsection{Purity constraints}
\label{sec:decl.func.purity}
\index{func purity constraints}

\begin{constraint}
  A \lain{func} shall not contain an unbounded \lain{while} loop (one
  without a \lain{decreasing} measure)~(\diagcode{E011}).
\end{constraint}

\begin{constraint}
  A \lain{func} shall not call a \lain{proc} or \lain{extern proc}~(\diagcode{E011}).
\end{constraint}

\begin{constraint}
  A \lain{func} shall not call itself directly or indirectly (mutual
  recursion)~(\diagcode{E011}).
\end{constraint}

\section{Procedure declarations}
\label{sec:decl.proc}
\index{procedure declaration}

\pnum
A procedure declaration introduces a \lain{proc}: a callable with
unrestricted side effects.  Its syntax is identical to \lain{func} except
for the \lain{proc} keyword.  A \lain{proc} may contain unbounded loops,
call other \lain{proc}s, and access global mutable state.

\pnum
The entry point of every Lain program shall be declared as:

\begin{laincode}
proc main() int { ... }
\end{laincode}

\begin{constraint}
  A program without a \lain{proc main() int} definition is
  \illformed{}~(\diagcode{E100}).
\end{constraint}

\section{Extern declarations}
\label{sec:decl.extern}
\index{extern declaration}

\pnum
An extern declaration introduces a C function into the Lain name space
without providing a body.  See \S\,\ref{sec:17-interop} for full semantics.

\begin{laincode}
extern proc printf(fmt *u8, ...) int
extern func strlen(s *u8) usize
\end{laincode}

\begin{constraint}
  Variadic parameters (\lain{...}) shall only appear in \lain{extern}
  declarations.  User-defined variadic functions are \illformed{}.
\end{constraint}

\section{Comptime declarations}
\label{sec:decl.comptime}
\index{comptime declaration}

\pnum
A top-level \lain{comptime \{ ... \}} block is evaluated by the compiler
before any runtime code is emitted.  It may define compile-time constants
or trigger \lain{compileError} assertions.
