% 12-functions.tex — Clause 12: Functions and procedures
\chapter{Functions and procedures}
\label{sec:12-functions}
\index{function}
\index{procedure}

\section{The func/proc distinction}
\label{sec:fn.distinction}
\index{func vs proc}

\pnum
Lain distinguishes between pure functions (\lain{func}) and side-effecting
procedures (\lain{proc}).  This distinction is enforced statically and
reflected in the type system.

\begin{longtable}{@{}lll@{}}
\toprule
\textbf{Property} & \textbf{\lain{func}} & \textbf{\lain{proc}} \\
\midrule
\endfirsthead
\toprule
\textbf{Property} & \textbf{\lain{func}} & \textbf{\lain{proc}} \\
\midrule
\endhead
\bottomrule
\endlastfoot
Side effects                      & forbidden          & allowed \\
Unbounded \lain{while}            & forbidden          & allowed \\
Bounded \lain{while decreasing}   & allowed            & allowed \\
\lain{for} loops                  & allowed            & allowed \\
Direct/indirect recursion         & forbidden          & allowed \\
Calling \lain{proc}               & forbidden          & allowed \\
Calling \lain{func}               & allowed            & allowed \\
Global mutable state              & no access          & full access \\
Termination guarantee             & yes                & no \\
\end{longtable}

\pnum
A \lain{func} is referentially transparent: calling it with the same
arguments always produces the same result.  Its evaluation has no
observable effect on any state outside its own stack frame.

\section{Purity enforcement}
\label{sec:fn.purity}
\index{purity enforcement}

\begin{constraint}
  A \lain{func} that calls a \lain{proc} or \lain{extern proc} is
  \illformed{}~(\diagcode{E011}).
\end{constraint}

\begin{constraint}
  A \lain{func} that contains an unbounded \lain{while} loop (one without
  a \lain{decreasing} measure) is \illformed{}~(\diagcode{E011}).
\end{constraint}

\begin{constraint}
  A \lain{func} that calls itself directly, or that participates in a
  mutual-recursion cycle, is \illformed{}~(\diagcode{E011}).  The
  implementation shall detect mutual recursion by performing a depth-first
  search on the call graph.
\end{constraint}

\begin{note}
  Generic functions (\lain{func} with \lain{comptime} parameters) are
  excluded from mutual-recursion detection until they are monomorphized.
\end{note}

\section{Termination analysis}
\label{sec:fn.termination}
\index{termination analysis}

\pnum
A \lain{func} is guaranteed to terminate because:
\begin{enumerate}
  \item all \lain{while} loops carry a verified termination measure;
  \item \lain{for} loops iterate over finite ranges;
  \item recursion is forbidden.
\end{enumerate}

\pnum
No termination analysis is performed for \lain{proc}s.

\section{Universal Function Call Syntax (UFCS)}
\label{sec:fn.ufcs}
\index{UFCS}

\pnum
An expression of the form \lain{x.f(args)} is desugared to \lain{f(x, args)}
before name resolution, provided \lain{f} is not a declared field of
\lain{x}'s type.  This is called Universal Function Call Syntax (UFCS).

\pnum
\textbf{UFCS ownership}: the ownership mode of the first argument in the
desugared call is determined by the declared mode of the first parameter of
\lain{f}:

\begin{itemize}
  \item If the first parameter is \lain{var}, the receiver is automatically
        mutably borrowed: \lain{x.f()} desugars to \lain{f(var x)}.
  \item If the first parameter is \lain{mov}, the receiver is automatically
        moved: \lain{x.f()} desugars to \lain{f(mov x)}.
  \item If the first parameter is shared (default), the receiver is shared:
        \lain{x.f()} desugars to \lain{f(x)}.
\end{itemize}

\pnum
\textbf{UFCS resolution algorithm}:
\begin{enumerate}
  \item Check if \lain{f} is a declared field of \lain{x}'s type.  If so,
        this is a field access, not UFCS.
  \item Search for a function named \lain{f} whose first parameter type is
        compatible with the type of \lain{x}.
  \item If found, desugar the call.
  \item If not found, emit \diagcode{E013} (undeclared identifier).
\end{enumerate}

\begin{example}
\begin{laincode}
func is_empty(s [u8]) bool { return s.len == 0 }

var buf = "hello"
var empty = buf.is_empty()   // desugars to: is_empty(buf)
\end{laincode}
\end{example}

\section{Calling conventions}
\label{sec:fn.abi}
\index{calling convention}

\pnum
The calling convention for Lain functions corresponds to the C99 calling
convention of the target platform, as determined by the C compiler used for
the translation output.  The parameter passing rules are:

\begin{longtable}{@{}lll@{}}
\toprule
\textbf{Lain parameter} & \textbf{C type} & \textbf{Semantics} \\
\midrule
\endfirsthead
\toprule
\textbf{Lain parameter} & \textbf{C type} & \textbf{Semantics} \\
\midrule
\endhead
\bottomrule
\endlastfoot
shared struct \lain{T} & \texttt{const T*} & pointer to read-only copy \\
\lain{var} struct \lain{T} & \texttt{T*}       & pointer, caller's copy \\
\lain{mov} struct \lain{T} & \texttt{T}        & by value (bitwise copy) \\
primitive any mode      & \texttt{T}        & by value \\
\end{longtable}

\section{Return values}
\label{sec:fn.return}
\index{return value}

\pnum
Return values follow the same rules as parameters for ownership modes.
A \lain{return mov x} transfers ownership of \lain{x} to the caller.

\begin{constraint}
  A \lain{return var} expression shall not refer to a local variable.
  It shall refer only to a field or element of a \lain{var} parameter
  or other data that outlives the function~(\diagcode{E010}).
\end{constraint}
