% 13-vra.tex — Clause 13: Value range constraints
\chapter{Value range constraints}
\label{sec:13-vra}
\index{value range analysis (VRA)}
\index{VRA}

\section{Overview}
\label{sec:vra.overview}

\pnum
Value Range Analysis (VRA) is a static analysis that tracks integer value
ranges through a program using interval arithmetic.  It is used to:
\begin{itemize}
  \item eliminate array bounds checks (\S\,\ref{sec:types.array});
  \item prove division-by-zero absence (\S\,\ref{sec:expr.arith});
  \item verify parameter and return constraints (\S\,\ref{sec:vra.constraints});
  \item prove in-guard safety (\S\,\ref{sec:expr.in}).
\end{itemize}

\pnum
VRA is decidable and polynomial in program size.  No SMT solver is required.
All analyses performed by a conforming implementation shall terminate.

\section{Range representation}
\label{sec:vra.range}
\index{range representation}

\pnum
A value range is an interval $[\ell,\, u]$ where $\ell \leq u$.  The
sentinels $-\infty$ and $+\infty$ represent the absence of a lower or upper
bound, respectively.  A range may additionally be \textit{unknown} (no
information available), in which case the interval is $(-\infty, +\infty)$.

\section{Range propagation rules}
\label{sec:vra.propagation}
\index{range propagation}

\pnum
The following table defines the result range for binary arithmetic
operations on intervals $[a,b]$ and $[c,d]$.  All arithmetic is
saturating: overflow wraps to $\pm\infty$.

\begin{longtable}{@{}lll@{}}
\toprule
\textbf{Operation} & \textbf{Result range} & \textbf{Notes} \\
\midrule
\endfirsthead
\toprule
\textbf{Operation} & \textbf{Result range} & \textbf{Notes} \\
\midrule
\endhead
\bottomrule
\endlastfoot
$[a,b] + [c,d]$ & $[a+c,\; b+d]$ & saturating \\
$[a,b] - [c,d]$ & $[a-d,\; b-c]$ & saturating \\
$[a,b] \times [c,d]$ & $[\min(ac,ad,bc,bd),\;\max(ac,ad,bc,bd)]$ & saturating \\
$[a,b] \div [c,d]$, $c > 0$ & $[a \div d,\; b \div c]$ & integer truncation \\
$[a,b] \div [c,d]$, $0 \in [c,d]$ & \diagcode{E015} & division by zero \\
$[a,b] \mathbin{\%} [c,d]$, $a \geq 0$, $c > 0$ & $[0,\; d-1]$ & \\
$[a,b] \mathbin{\%} [c,d]$, otherwise & unknown & conservative \\
\end{longtable}

\section{Conditional refinement}
\label{sec:vra.refinement}
\index{conditional refinement}

\pnum
After an \lain{if x < N} test, the VRA refines the range of \lain{x}
in the then-branch to $[\ell,\, N-1]$ and in the else-branch to $[N,\, u]$
(where $[\ell, u]$ is the range before the test).

\pnum
At the join point after an \lain{if}/\lain{else}, the range is the union
of the ranges from both branches.

\pnum
When conditions are chained with \lain{and}, refinements from the
left operand are active when evaluating the right operand
(\S\,\ref{sec:expr.in}).

\section{In-guard semantics}
\label{sec:vra.in-guard}
\index{in guard!VRA semantics}

\pnum
An in-guard \lain{idx in arr} (\S\,\ref{sec:expr.in}) introduces the
constraint $0 \leq \mathit{idx} < \mathit{arr.len}$ into the VRA's
range environment.  Inside the guarded body, \lain{arr[idx]} is accepted
without further proof.

\pnum
When \lain{idx in arr} is the condition of a \lain{while decreasing M}
loop, the VRA infers that $\mathit{arr.len} - \mathit{idx} \geq 0$ (a
valid non-negativity proof for the measure $\mathit{arr.len} - \mathit{idx}$).

\section{Constraint annotations}
\label{sec:vra.constraints}
\index{constraint annotation}

\pnum
A constraint annotation of the form \lain{T op bound} on a parameter or
return type declares an invariant that the VRA shall enforce:

\begin{itemize}
  \item \textbf{Parameter constraint}: at every call site, the VRA shall
        prove that the argument satisfies the constraint.
  \item \textbf{Return constraint}: at every \lain{return} statement, the
        VRA shall prove that the return expression satisfies the constraint.
\end{itemize}

\begin{constraint}
  A function call where the argument range cannot be proven to satisfy
  a parameter constraint is \illformed{}~(\diagcode{E012}).
\end{constraint}

\begin{constraint}
  A \lain{return} expression whose range cannot be proven to satisfy the
  declared return constraint is \illformed{}~(\diagcode{E012}).
\end{constraint}

\section{Interprocedural range propagation}
\label{sec:vra.interprocedural}
\index{interprocedural VRA}

\pnum
When a call expression appears in a context where the VRA needs a range,
it uses the callee's declared return constraint (if any) as a conservative
approximation.

\begin{example}
\begin{laincode}
func nonzero() int >= 1 { return 42 }

func divide(a int) int {
    return a / nonzero()   // divisor range known >= 1: no [E015]
}
\end{laincode}
\end{example}

\pnum
When the callee has no declared return constraint, the call expression has
an unknown range.

\begin{note}
  Interprocedural VRA is conservative: only declared constraints are used,
  not body analysis.  This preserves decidability and ensures that changing
  a function's body cannot silently affect its callers' safety proofs.
\end{note}

\section{Limits and known gaps}
\label{sec:vra.limits}

\pnum
The following limitations are known and accepted:

\begin{enumerate}
  \item VRA does not track ranges of struct fields.  A struct field accessed
        through a pointer always has an unknown range.
  \item VRA does not track ranges across loop-carried dependences with
        non-linear measures.
  \item Integer overflow in VRA arithmetic uses saturating semantics; the
        actual Lain program uses wrapping arithmetic.  Programs whose
        correctness depends on integer overflow are not analyzed correctly
        by VRA.
  \item Floating-point values are not tracked by VRA.
\end{enumerate}
