% 15-match.tex — Clause 15: Pattern matching
\chapter{Pattern matching}
\label{sec:15-match}
\index{pattern matching}

\section{Overview}
\label{sec:match.overview}

\pnum
The \lain{case} expression matches a value against a sequence of patterns
and evaluates the body of the first matching arm.  Every \lain{case}
expression shall be exhaustive: the patterns shall collectively cover all
possible values of the scrutinee type.  The grammar production is
\gramref{match-expression} (Annex~A).

\section{Syntax}
\label{sec:match.syntax}

\begin{laincode}
case expr {
    Pattern1: body1
    Pattern2: body2
    else:     default_body
}
\end{laincode}

\pnum
Each arm consists of a pattern (or comma-separated list of patterns) and a
body.  The body may be a single expression or a block.  The arms are
tested in declaration order; the first matching arm is taken.

\section{Patterns}
\label{sec:match.patterns}

\pnum
The following pattern forms are supported:

\begin{description}[leftmargin=2em]
  \item[Wildcard (\lain{else})]
    Matches any value.  Must be the last arm.  Required for integer and
    character scrutinees.

  \item[Literal pattern]
    Matches a specific integer or character value:
    \lain{42:}, \lain{'a':}.

  \item[Enum/ADT variant pattern]
    Matches a specific variant.  If the variant has fields, they are bound:
    \lain{Circle(r): ...} binds the radius to \lain{r}.

  \item[Range pattern]
    Matches an inclusive range of integer or character values:
    \lain{1..10:} matches 1 through 10.

  \item[Multi-pattern (OR)]
    A comma-separated list of patterns on a single arm:
    \lain{1, 2, 3: ...}.
\end{description}

\section{Exhaustiveness}
\label{sec:match.exhaustiveness}
\index{exhaustiveness}

\pnum
A \lain{case} expression is exhaustive if the set of patterns covers all
values of the scrutinee type.  The rules by type are:

\begin{longtable}{@{}ll@{}}
\toprule
\textbf{Scrutinee type} & \textbf{Exhaustiveness requirement} \\
\midrule
\endfirsthead
\toprule
\textbf{Scrutinee type} & \textbf{Exhaustiveness requirement} \\
\midrule
\endhead
\bottomrule
\endlastfoot
Enum     & All variants covered, or \lain{else} arm present \\
ADT      & All variants covered, or \lain{else} arm present \\
Integer  & \lain{else} arm always required \\
\lain{u8} (char) & \lain{else} arm always required \\
\lain{bool} & \lain{true} and \lain{false} covered, or \lain{else} \\
\end{longtable}

\begin{constraint}
  A non-exhaustive \lain{case} expression is
  \illformed{}~(\diagcode{E014}).
\end{constraint}

\begin{example}
\begin{laincode}
enum Color { Red, Green, Blue }
var c = Color.Red

case c {
    Red:   return 1
    Green: return 2
    Blue:  return 3
}  // exhaustive: all three variants covered
\end{laincode}
\end{example}

\section{Ownership and match}
\label{sec:match.ownership}
\index{ownership in match}

\pnum
By default, \lain{case expr} may consume the scrutinee if it is a linear
type.  Bindings introduced by variant patterns in consuming arms hold
ownership of the bound values.

\pnum
\lain{case \&expr} performs a non-consuming match: the scrutinee is not
consumed but instead temporarily borrowed for the duration of the match
body (see \S\,\ref{sec:own.borrowed-match}).

\begin{constraint}
  Inside a \lain{case \&expr} body, the scrutinee shall not be mutated or
  moved~(\diagcode{E004}).
\end{constraint}

\section{Case expression result type}
\label{sec:match.type}

\pnum
When \lain{case} is used as an expression, all arms shall produce values
of compatible types (\S\,\ref{sec:types.compat}).  The type of the
\lain{case} expression is the common type.

\begin{constraint}
  Arms of a \lain{case} expression that produce incompatible types are
  \illformed{}.
\end{constraint}

\section{String pattern matching}
\label{sec:match.string}
\index{string pattern matching}

\pnum
String literals may appear as patterns.  A string pattern matches if the
scrutinee has the same byte sequence and length as the pattern literal.
The comparison is performed byte-by-byte (analogous to \texttt{memcmp}).

\begin{note}
  String pattern matching in the translation output is implemented as a
  sequence of byte comparisons or \texttt{memcmp} calls, depending on
  the implementation.  Its asymptotic cost is O(n) in the pattern length.
\end{note}
