% 16-modules.tex — Clause 16: Module system
\chapter{Module system}
\label{sec:16-modules}
\index{module system}

\section{Overview}
\label{sec:mod.overview}

\pnum
Each Lain source file defines one module.  Modules are the unit of code
organization and compilation.  All names declared at the top level of a
source file are in that module's scope and are accessible from other
modules that import the module.

\section{Module declarations}
\label{sec:mod.declaration}
\index{module declaration}

\pnum
A top-level \lain{module path.name} declaration declares the module name
for the current source file.  Module declarations are optional; if absent,
the module name is \idbehav{derived from the file path by the implementation}.

\section{Import declarations}
\label{sec:mod.import}
\index{import declaration}

\pnum
\lain{import module.path} makes the declarations of the named module
available in the current source file.  The grammar production is
\gramref{import-declaration} (Annex~A).

\pnum
\textbf{Path resolution}: a module path is converted to a file system path
by replacing dots with directory separators and appending \texttt{.ln}.
The search is relative to the compiler's working directory.
\idbehav{The exact search algorithm, including support for search paths,
is implementation-defined}.

\begin{longtable}{@{}ll@{}}
\toprule
\textbf{Module path} & \textbf{File path} \\
\midrule
\endfirsthead
\toprule
\textbf{Module path} & \textbf{File path} \\
\midrule
\endhead
\bottomrule
\endlastfoot
\lain{std.c}        & \texttt{std/c.ln}        \\
\lain{std.io}       & \texttt{std/io.ln}        \\
\lain{std.fs}       & \texttt{std/fs.ln}        \\
\lain{std.math}     & \texttt{std/math.ln}      \\
\lain{mylib.utils}  & \texttt{mylib/utils.ln}   \\
\end{longtable}

\section{Module loading}
\label{sec:mod.loading}
\index{module loading}

\pnum
When the compiler encounters an \lain{import} declaration:
\begin{enumerate}
  \item The module path is converted to a file path.
  \item If the module has already been loaded, the import is a no-op
        (deduplication).
  \item Otherwise, the file is read, lexed, parsed, and recursively
        resolved.
  \item The loaded module's declarations are merged into the importing
        module's declaration list.
\end{enumerate}

\pnum
Each module is loaded at most once.  Circular imports are therefore
implicitly handled: if module~A imports module~B which imports module~A,
the second import of~A is a no-op because A is already loaded.

\begin{note}
  The reference implementation maintains a linked list of loaded modules
  keyed by their dot-separated name.
\end{note}

\section{Name visibility}
\label{sec:mod.visibility}
\index{name visibility}

\pnum
All names declared at module scope are accessible to any module that
imports the declaring module.  There is no private/public distinction in
the current version of the language.
\idbehav{Future versions may add visibility modifiers (e.g., \texttt{export})}.

\section{Standard library module paths}
\label{sec:mod.stdlib}
\index{standard library!module paths}

\pnum
The following module paths are defined by this specification for the
standard library:

\begin{longtable}{@{}ll@{}}
\toprule
\textbf{Module path} & \textbf{Description} \\
\midrule
\endfirsthead
\toprule
\textbf{Module path} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endlastfoot
\lain{std.c}    & C standard library bindings \\
\lain{std.io}   & Console I/O \\
\lain{std.fs}   & File system operations \\
\lain{std.math} & Mathematical functions \\
\end{longtable}

\section{Name mangling}
\label{sec:mod.mangling}
\index{name mangling}

\pnum
The name of a Lain declaration in the translation output is
\idbehav{determined by the implementation}.  The reference implementation
uses the function name prefixed with the module path, with dots replaced
by underscores.  Name collisions between modules are
\idbehav{detected by the implementation; the reference implementation
appends a numeric suffix to resolve conflicts}.
