%% B-diagnostics.tex — Annex B: Diagnostic codes (normative)
\chapter{Diagnostic codes}
\label{annex:B-diagnostics}
\index{diagnostic codes}

\pnum
This annex is normative.  It lists every diagnostic code defined by this
specification, the clause that requires the diagnostic, the category
(constraint violation, warning, or informational), and a representative
message text.  A conforming implementation shall issue a diagnostic for
every constraint violation in the table below.

\pnum
Diagnostic codes are divided into the following ranges:

\begin{longtable}{@{}ll@{}}
\toprule
\textbf{Range} & \textbf{Domain} \\
\midrule
\endfirsthead
\toprule
\textbf{Range} & \textbf{Domain} \\
\midrule
\endhead
\bottomrule
\endlastfoot
\diagcode{E001}–\diagcode{E010} & Ownership and lifetime \\
\diagcode{E011}–\diagcode{E015} & Purity, safety, and verification \\
\diagcode{E045}                 & Compile-time evaluation \\
\diagcode{E080}–\diagcode{E082} & Bounded while verification \\
\diagcode{E100}                 & Miscellaneous \\
\diagcode{VRA}                  & Value range analysis \\
\end{longtable}

% -------------------------------------------------------------------------
\section*{Ownership and lifetime diagnostics (\diagcode{E001}–\diagcode{E010})}
% -------------------------------------------------------------------------

\begin{longtable}{@{}p{1.5cm}p{3.5cm}p{7.5cm}@{}}
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endfirsthead
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endhead
\bottomrule
\endlastfoot

\diagcode{E001}
  & \S\,\ref{sec:own.use-after-move}
  & \textit{use of moved value '\texttt{x}'} ---
    Issued when an expression reads or passes a variable whose linear
    state is \textsc{Consumed}.
    \\[4pt]

\diagcode{E002}
  & \S\,\ref{sec:own.use-after-move}
  & \textit{value '\texttt{x}' consumed twice} ---
    Issued when a second \lain{mov} expression is applied to a variable
    already in state \textsc{Consumed}.
    \\[4pt]

\diagcode{E003}
  & \S\,\ref{sec:own.linear-must-consume}
  & \textit{linear value '\texttt{x}' not consumed before end of scope} ---
    Issued when a \lain{mov}-mode variable exits scope in state
    \textsc{Unconsumed}.
    \\[4pt]

\diagcode{E004}
  & \S\,\ref{sec:own.borrow-conflict}
  & \textit{cannot borrow '\texttt{x}' as mutable; it is already borrowed}
    --- Issued when a mutable borrow conflicts with an existing borrow
    (reads or write).
    \\[4pt]

\diagcode{E005}
  & \S\,\ref{sec:own.borrow-conflict}
  & \textit{cannot borrow '\texttt{x}' as shared; it is already mutably
    borrowed} --- Issued when a shared borrow is requested while a
    mutable borrow is active.
    \\[4pt]

\diagcode{E006}
  & \S\,\ref{sec:own.copy-linear}
  & \textit{cannot copy linear type '\texttt{T}'} ---
    Issued when assignment or argument passing implicitly copies a type
    that has a \lain{mov}-mode field.
    \\[4pt]

\diagcode{E007}
  & \S\,\ref{sec:own.modes}
  & \textit{cannot assign through a shared reference '\texttt{x}'} ---
    Issued when a write is attempted through a shared (non-\lain{var})
    binding.
    \\[4pt]

\diagcode{E010}
  & \S\,\ref{sec:mem.storage}
  & \textit{cannot return reference to local variable '\texttt{x}'} ---
    Issued when a \lain{return} statement returns a reference whose
    referent has automatic storage duration.
    \\[4pt]

\end{longtable}

% -------------------------------------------------------------------------
\section*{Purity, safety, and verification (\diagcode{E011}–\diagcode{E015})}
% -------------------------------------------------------------------------

\begin{longtable}{@{}p{1.5cm}p{3.5cm}p{7.5cm}@{}}
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endfirsthead
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endhead
\bottomrule
\endlastfoot

\diagcode{E011}
  & \S\,\ref{sec:12-functions.purity}
  & \textit{impure operation in pure function '\texttt{f}'} ---
    Issued when a \lain{func} body contains a call to a \lain{proc}, an
    unbounded \lain{while} without a \lain{decreasing} measure, or a
    direct/mutual recursion without a measure.
    \\[4pt]

\diagcode{E012}
  & \S\,\ref{sec:interop.pointers}
  & \textit{pointer cast outside unsafe block} ---
    Issued when a pointer-to-pointer or integer-to-pointer cast occurs
    outside an \lain{unsafe} block.
    \\[4pt]

\diagcode{E013}
  & \S\,\ref{sec:06-basics.decl}
  & \textit{use of undeclared identifier '\texttt{x}'} ---
    Issued when an identifier is referenced before its declaration in
    scope.
    \\[4pt]

\diagcode{E014}
  & \S\,\ref{sec:08-expressions.index},
    \S\,\ref{sec:13-vra}
  & \textit{index '\texttt{i}' not proven in bounds of '\texttt{arr}'} ---
    Issued when VRA cannot prove that an array index expression is
    within $[0,\, n-1]$.  Also issued for non-exhaustive \lain{case}
    expressions when combined with the exhaustiveness diagnostic (shared
    code path).
    \\[4pt]

\diagcode{E015}
  & \S\,\ref{sec:08-expressions.arith}
  & \textit{division by zero} ---
    Issued when VRA can prove statically that a division or modulo
    expression has a zero divisor.
    \\[4pt]

\end{longtable}

% -------------------------------------------------------------------------
\section*{Compile-time evaluation (\diagcode{E045})}
% -------------------------------------------------------------------------

\begin{longtable}{@{}p{1.5cm}p{3.5cm}p{7.5cm}@{}}
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endfirsthead
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endhead
\bottomrule
\endlastfoot

\diagcode{E045}
  & \S\,\ref{sec:gen.recursion-depth}
  & \textit{comptime evaluation depth limit exceeded} ---
    Issued when compile-time function evaluation recurses beyond the
    implementation's depth limit (256 in the reference implementation).
    \\[4pt]

\end{longtable}

% -------------------------------------------------------------------------
\section*{Bounded while verification (\diagcode{E080}–\diagcode{E082})}
% -------------------------------------------------------------------------

\begin{longtable}{@{}p{1.5cm}p{3.5cm}p{7.5cm}@{}}
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endfirsthead
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endhead
\bottomrule
\endlastfoot

\diagcode{E080}
  & \S\,\ref{sec:stmt.while.bounded}
  & \textit{termination measure not proven non-negative} ---
    Issued when the compiler cannot prove that the \lain{decreasing}
    expression is $\geq 0$ whenever the loop condition holds.
    \\[4pt]

\diagcode{E081}
  & \S\,\ref{sec:stmt.while.bounded}
  & \textit{termination measure does not depend on a loop variable} ---
    Issued when the \lain{decreasing} expression consists entirely of
    compile-time constants and therefore cannot strictly decrease.
    \\[4pt]

\diagcode{E082}
  & \S\,\ref{sec:stmt.while.bounded}
  & \textit{termination measure not proven to decrease in loop body} ---
    Issued when the compiler cannot verify that every execution of the
    loop body strictly decreases the measure by at least 1.
    \\[4pt]

\end{longtable}

% -------------------------------------------------------------------------
\section*{Miscellaneous (\diagcode{E100})}
% -------------------------------------------------------------------------

\begin{longtable}{@{}p{1.5cm}p{3.5cm}p{7.5cm}@{}}
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endfirsthead
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endhead
\bottomrule
\endlastfoot

\diagcode{E100}
  & \S\,\ref{sec:05-lexical}, \S\,\ref{sec:07-types}
  & \textit{ill-formed token / semantic constraint violation} ---
    A catch-all code used for lexical and syntactic errors without a
    dedicated code, including: missing semicolons (\S\,\ref{sec:lex.semi}),
    using \lain{==} on enum types (\S\,\ref{sec:types.enum}), and other
    single-token syntactic errors.
    \\[4pt]

\end{longtable}

% -------------------------------------------------------------------------
\section*{Value range analysis (\diagcode{VRA})}
% -------------------------------------------------------------------------

\begin{longtable}{@{}p{1.5cm}p{3.5cm}p{7.5cm}@{}}
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endfirsthead
\toprule
\textbf{Code} & \textbf{Clause} & \textbf{Message and trigger} \\
\midrule
\endhead
\bottomrule
\endlastfoot

\diagcode{VRA}
  & \S\,\ref{sec:13-vra}
  & \textit{value range proof required} ---
    Issued when VRA cannot prove that an operation satisfies its
    required constraint and more specific code is not available.
    Appears as a diagnostic prefix in messages such as:
    ``\texttt{[VRA] cannot prove index in bounds}'' and
    ``\texttt{[VRA] return value may not satisfy constraint}''.
    \\[4pt]

\end{longtable}
