%% C-rationale.tex — Annex C: Language design rationale (informative)
\chapter{Language design rationale}
\label{annex:C-rationale}
\index{rationale}

\pnum
This annex is informative.  It explains the reasoning behind key design
decisions in the Lain language.  Understanding the rationale helps readers
interpret edge cases in the normative text and guides implementors when
the specification is ambiguous.

% =========================================================================
\section{The \texttt{func}/\texttt{proc} distinction}
\label{sec:rationale.func-proc}
\index{func/proc distinction!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Separate pure functions (\lain{func}) from
side-effecting procedures (\lain{proc}) at the declaration level.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{Termination guarantees}: a \lain{func} is guaranteed to
        terminate (no unbounded \lain{while}, no unchecked recursion).
        This enables the compiler to reason statically about function
        completion and to use \lain{func} bodies inside \lain{comptime}
        expressions.
  \item \textit{Purity propagation}: the type system propagates purity
        constraints — a \lain{func} cannot call a \lain{proc}, so pure
        code stays pure by construction.
  \item \textit{Optimisation}: the compiler may freely reorder, memoize,
        or eliminate calls to \lain{func} because they have no observable
        side effects.
  \item \textit{Readability}: the declaration keyword tells the reader
        whether a callable has side effects without inspecting its body.
\end{itemize}

\pnum
\textbf{Alternatives considered.}
Single keyword with a purity annotation (\lain{func(pure)}) was rejected
as less visible.  A Haskell-style IO monad was rejected as too complex
for a systems language.

% =========================================================================
\section{Explicit \texttt{mov} at call sites}
\label{sec:rationale.mov}
\index{mov!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Require explicit \lain{mov} at call sites when
transferring ownership.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{Readability}: \lain{close\_file(mov f)} makes ownership
        transfer visible at the call site.
  \item \textit{Intentionality}: accidental ownership transfer is a
        common source of bugs.  Requiring \lain{mov} forces the
        programmer to acknowledge the transfer.
  \item \textit{Symmetry}: the annotation appears in both declarations
        (\lain{mov x T}) and at call sites (\lain{mov x}), creating a
        uniform language.
\end{itemize}

\pnum
\textbf{Alternatives considered.}
Rust-style implicit moves were rejected because they are invisible at
call sites.  A library function \texttt{std::move} (C++) was rejected
because ownership transfer should be a language primitive, not a library
function.

% =========================================================================
\section{Implicit NLL lifetimes (no annotations)}
\label{sec:rationale.nll}
\index{lifetime annotations!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Borrows have implicit lifetimes determined by
non-lexical lifetime analysis.  No explicit lifetime annotations are
required.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{Simplicity}: Rust's lifetime annotations are widely
        cited as its steepest learning curve.  Lain targets programmers
        who want safety without that complexity.
  \item \textit{Sufficient expressiveness}: NLL handles the vast
        majority of real-world borrow patterns.
  \item \textit{Predictability}: borrows expire at last use — a single,
        intuitive rule.
\end{itemize}

\pnum
\textbf{Trade-offs.}  Some programs that Rust can verify with explicit
lifetime annotations may be rejected by Lain's more conservative
analysis.  Higher-order borrowing patterns are constrained.

% =========================================================================
\section{C99 translation output}
\label{sec:rationale.c99}
\index{C99 backend!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Compile to C99 source code rather than LLVM IR,
machine code, or bytecode.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{Portability}: C99 compilers exist for every platform,
        from server-class x86 to microcontrollers.  Lain inherits this
        portability for free.
  \item \textit{Toolchain reuse}: users get GCC/Clang optimisation passes,
        debuggers, profilers, and sanitisers without custom tooling.
  \item \textit{Interoperability}: C interop is zero-cost because the
        output \emph{is} C.
  \item \textit{Simplicity}: a C emitter is orders of magnitude simpler
        than an LLVM backend or a native code generator.
  \item \textit{Auditability}: the generated C code is human-readable.
\end{itemize}

\pnum
\textbf{Trade-offs.}  Compilation is two-phase (Lain~→~C~→~binary).
Some optimisations are impossible without direct IR control.  The C type
system constrains what Lain can express (e.g., no guaranteed tail calls).

% =========================================================================
\section{No exception mechanism}
\label{sec:rationale.no-exceptions}
\index{exceptions!absence of, rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Lain has no exception mechanism.  All error handling
uses return values.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{Zero overhead}: no unwinding tables, catch tables, or
        RTTI.
  \item \textit{Determinism}: error handling follows the same control
        flow as normal code; no invisible jumps.
  \item \textit{Embedded compatibility}: stack unwinding is unavailable
        or unreliable on many embedded platforms.
  \item \textit{Explicit error paths}: every error is visible in the
        function signature (via \lain{Result} return types).
  \item \textit{Compatibility with \lain{defer}}: the \lain{defer}
        mechanism provides deterministic cleanup without exceptions.
\end{itemize}

% =========================================================================
\section{Explicit \texttt{unsafe} blocks}
\label{sec:rationale.unsafe}
\index{unsafe!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Pointer dereference and direct ADT field access
require an explicit \lain{unsafe} block.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{Auditability}: safety-critical codebases can grep for
        \lain{unsafe \{} to enumerate all potentially dangerous
        operations.
  \item \textit{Minimal scope}: Lain's \lain{unsafe} is narrow — it
        enables only pointer dereference, address-of, casts, direct ADT
        access, and bounds suppression.  Unlike C (where everything is
        unsafe by default) or some languages (where \lain{unsafe} enables
        a much larger capability set), Lain's \lain{unsafe} is a
        surgical escape hatch.
  \item \textit{Social contract}: the keyword communicates to code
        reviewers that extra scrutiny is warranted.
\end{itemize}

% =========================================================================
\section{Value Range Analysis instead of SMT}
\label{sec:rationale.vra}
\index{VRA!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Use interval-based VRA for static verification
instead of an SMT solver.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{Decidability}: VRA always terminates in polynomial time.
        SMT solving is NP-hard in the general case and may time out.
  \item \textit{Predictability}: programmers can reason about what the
        compiler will accept by thinking about integer ranges.  SMT
        solver reasoning is opaque.
  \item \textit{Simplicity}: the VRA implementation is approximately 600 lines.
        An SMT integration would be orders of magnitude more complex and
        would introduce an external dependency.
\end{itemize}

\pnum
\textbf{Trade-offs.}  VRA is incomplete — it may reject valid programs
that an SMT solver could prove correct.  Non-linear constraints produce
conservatively wide ranges.

% =========================================================================
\section{Comptime generics instead of \texttt{<T>} syntax}
\label{sec:rationale.comptime}
\index{generics!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Use compile-time function evaluation with
\lain{comptime} parameters instead of traditional generic syntax.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{Unification}: generics, type aliases, and compile-time
        assertions all use the same mechanism.
  \item \textit{Expressiveness}: comptime functions can contain arbitrary
        logic, enabling type-level programming that traditional generics
        cannot express.
  \item \textit{Simplicity}: no angle brackets, no trait bounds, no
        where clauses.  Just functions that take types and return types.
  \item \textit{Precedent}: inspired by Zig's comptime approach, which
        has proven ergonomic in practice.
\end{itemize}

% =========================================================================
\section{Universal Function Call Syntax (UFCS)}
\label{sec:rationale.ufcs}
\index{UFCS!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Support UFCS: \lain{x.f(args)} desugars to
\lain{f(x, args)}.

\pnum
\textbf{Rationale.}
\begin{itemize}
  \item \textit{No classes needed}: UFCS provides method-like syntax
        without class hierarchies, virtual dispatch, or inheritance.
  \item \textit{Chainability}: enables fluent APIs.
  \item \textit{Discoverability}: IDE auto-completion works naturally —
        typing a dot after a value shows available operations.
  \item \textit{Simplicity}: syntactic sugar only; no new concepts, no
        method tables, no implicit \lain{this} pointer.
\end{itemize}

% =========================================================================
\section{Bounded \texttt{while} in \texttt{func}}
\label{sec:rationale.bounded-while}
\index{bounded while!rationale}
% =========================================================================

\pnum
\textbf{Decision.}  Allow \lain{while} loops in \lain{func} bodies when
a \lain{decreasing} termination measure is supplied.

\pnum
\textbf{Rationale.}
Without \lain{decreasing}, any \lain{while} loop in a \lain{func} would
require a proof of termination that goes beyond what the compiler can
verify automatically.  Requiring the programmer to supply a
\textit{measure} — a non-negative integer expression that strictly
decreases on every iteration — is a well-established technique from
program verification (\textit{ranking functions}).  The compiler's three
checks (non-negativity, variable dependency, strict decrease) are
sufficient to verify termination without an external solver.

\pnum
This allows algorithms such as binary search and GCD computation to be
expressed as \lain{func} without resorting to \lain{proc}.
