%% D-comparison.tex — Annex D: Comparison with related languages (informative)
\chapter{Comparison with related languages}
\label{annex:D-comparison}
\index{language comparison}

\pnum
This annex is informative.  It places Lain in the context of other
systems-programming languages to help readers understand its design
choices.  Comparisons are accurate as of the date of this specification;
language features evolve and readers should verify current status.

% =========================================================================
\section{Feature matrix}
\label{sec:cmp.matrix}
% =========================================================================

\begin{longtable}{@{}p{4cm}p{1.5cm}p{1.5cm}p{1.5cm}p{1.5cm}p{1.5cm}@{}}
\toprule
\textbf{Feature} & \textbf{Lain} & \textbf{Rust} & \textbf{C} & \textbf{Zig} & \textbf{Go} \\
\midrule
\endfirsthead
\toprule
\textbf{Feature} & \textbf{Lain} & \textbf{Rust} & \textbf{C} & \textbf{Zig} & \textbf{Go} \\
\midrule
\endhead
\bottomrule
\endlastfoot
Memory safety       & Compile-time & Compile-time & Manual & Runtime+CT & GC \\
Null safety         & Option type  & Option type  & None   & Optional   & Nil \\
Exceptions          & None (Result)& None (Result)& errno  & Error union& panic/recover \\
Generics            & Comptime     & Trait-based  & Macros & Comptime   & Type params \\
Lifetime annots.    & None (NLL)   & Required     & N/A    & None       & N/A \\
Purity tracking     & \lain{func}/\lain{proc} & None & None & None & None \\
Pattern matching    & Exhaustive   & Exhaustive   & switch & switch     & switch \\
UFCS                & Yes          & No           & No     & No         & No \\
Backend             & C99          & LLVM         & Native & LLVM       & Go compiler \\
GC / runtime        & None         & Minimal      & None   & Minimal    & GC \\
\end{longtable}

% =========================================================================
\section{Lain versus Rust}
\label{sec:cmp.rust}
\index{Rust comparison}
% =========================================================================

\subsection*{Similarities}
\begin{itemize}
  \item Ownership-based memory safety with no GC.
  \item Move semantics with explicit transfer.
  \item No null pointers (Option type for optional values).
  \item Pattern matching with exhaustiveness checking.
  \item No exceptions; Result type for error propagation.
  \item Zero-cost abstractions.
\end{itemize}

\subsection*{Differences}

\begin{longtable}{@{}p{4.5cm}p{5.5cm}p{5cm}@{}}
\toprule
\textbf{Aspect} & \textbf{Lain} & \textbf{Rust} \\
\midrule
\endfirsthead
\toprule
\textbf{Aspect} & \textbf{Lain} & \textbf{Rust} \\
\midrule
\endhead
\bottomrule
\endlastfoot
Lifetime annotations  & Implicit (NLL only) & Explicit (\texttt{'a}, \texttt{'static}) \\
Mutable borrow syntax & \lain{var x} at call site & \texttt{\&mut x} \\
Move syntax           & Explicit \lain{mov x} & Implicit (default) \\
Generics              & Comptime functions & Trait-based + monomorphization \\
Purity                & \lain{func}/\lain{proc} enforced & Convention only \\
Async                 & Not supported & \texttt{async}/\texttt{await} \\
Traits / Interfaces   & Not supported & Full trait system \\
Macros                & Not supported & \texttt{macro\_rules!} + proc macros \\
Backend               & C99 & LLVM \\
Learning curve        & Lower & Higher \\
\end{longtable}

\pnum
\textbf{Key insight.}  Lain trades Rust's expressiveness (explicit lifetime
annotations, trait system, async/await) for simplicity.  The result is
a language that is easier to learn but less expressive for complex
borrowing patterns.

% =========================================================================
\section{Lain versus C}
\label{sec:cmp.c}
\index{C comparison}
% =========================================================================

\subsection*{Similarities}
\begin{itemize}
  \item Compiles to native code; no GC.
  \item Explicit memory management.
  \item No runtime overhead.
  \item Systems-level programming: raw pointers accessible in \lain{unsafe}.
\end{itemize}

\subsection*{Differences}

\begin{longtable}{@{}p{4.5cm}p{5.5cm}p{5cm}@{}}
\toprule
\textbf{Aspect} & \textbf{Lain} & \textbf{C} \\
\midrule
\endfirsthead
\toprule
\textbf{Aspect} & \textbf{Lain} & \textbf{C} \\
\midrule
\endhead
\bottomrule
\endlastfoot
Memory safety         & Compile-time guarantees & Manual (error-prone) \\
Type safety           & Strong, narrow widening & Weak, implicit conversion \\
Ownership             & Tracked by compiler     & Programmer discipline \\
Bounds checking       & Static (VRA)            & None \\
Division by zero      & Static detection (E015) & Undefined behavior \\
Pattern matching      & Exhaustive \lain{case}  & \texttt{switch} (fall-through) \\
Null safety           & Option type             & Null everywhere \\
Strings               & Slices with length      & Null-terminated arrays \\
Generics              & Comptime functions      & Macros / \texttt{void*} \\
Header files          & Module system           & \texttt{\#include} \\
\end{longtable}

\pnum
\textbf{Key insight.}  Lain provides C-level performance with compile-time
safety guarantees.  The C99 backend enables seamless interop with existing
C libraries while eliminating the most common classes of C undefined
behavior from the safe fragment.

% =========================================================================
\section{Lain versus Zig}
\label{sec:cmp.zig}
\index{Zig comparison}
% =========================================================================

\subsection*{Similarities}
\begin{itemize}
  \item Comptime-based generics.
  \item No hidden control flow; no hidden allocations.
  \item C interoperability as a first-class concern.
  \item Explicit error handling with no exceptions.
\end{itemize}

\subsection*{Differences}

\begin{longtable}{@{}p{4.5cm}p{5.5cm}p{5cm}@{}}
\toprule
\textbf{Aspect} & \textbf{Lain} & \textbf{Zig} \\
\midrule
\endfirsthead
\toprule
\textbf{Aspect} & \textbf{Lain} & \textbf{Zig} \\
\midrule
\endhead
\bottomrule
\endlastfoot
Memory safety         & Compile-time ownership  & Runtime (debug) \\
Ownership system      & Borrow checker          & Manual \\
Backend               & C99                     & LLVM + self-hosted \\
Comptime scope        & Type-parameter functions & Full expression evaluation \\
Error handling        & Result type + \lain{case} & Error unions + \texttt{try} \\
Allocator model       & C interop (malloc/free) & Custom allocators everywhere \\
\end{longtable}

\pnum
\textbf{Key insight.}  Both Lain and Zig use comptime for generics.
Zig's comptime is more powerful (arbitrary expression evaluation);
Lain's ownership system provides stronger compile-time memory safety
guarantees than Zig's debug-mode runtime checks.

% =========================================================================
\section{Lain versus Go}
\label{sec:cmp.go}
\index{Go comparison}
% =========================================================================

\subsection*{Similarities}
\begin{itemize}
  \item Simple, readable syntax.
  \item Fast compilation.
  \item No class hierarchies.
\end{itemize}

\subsection*{Differences}

\begin{longtable}{@{}p{4.5cm}p{5.5cm}p{5cm}@{}}
\toprule
\textbf{Aspect} & \textbf{Lain} & \textbf{Go} \\
\midrule
\endfirsthead
\toprule
\textbf{Aspect} & \textbf{Lain} & \textbf{Go} \\
\midrule
\endhead
\bottomrule
\endlastfoot
Memory management     & Ownership (no GC)       & Garbage collected \\
Runtime               & None                    & GC + goroutine scheduler \\
Generics              & Comptime functions      & Type parameters \\
Error handling        & Result type             & Multiple return values \\
Null safety           & Option type             & Nil (null) \\
Concurrency           & Not supported           & Goroutines + channels \\
Determinism           & Yes (no GC pauses)      & GC pauses possible \\
\end{longtable}

\pnum
\textbf{Key insight.}  Go achieves simplicity through garbage collection.
Lain achieves similar simplicity goals through ownership-based safety,
without a runtime — making Lain suitable for embedded and real-time
applications where Go's GC is problematic.

% =========================================================================
\section{Features unique to Lain}
\label{sec:cmp.unique}
% =========================================================================

\pnum
The following features distinguish Lain from all languages in the
comparison:

\begin{longtable}{@{}p{4.5cm}p{9cm}@{}}
\toprule
\textbf{Feature} & \textbf{Description} \\
\midrule
\endfirsthead
\toprule
\textbf{Feature} & \textbf{Description} \\
\midrule
\endhead
\bottomrule
\endlastfoot
\lain{func}/\lain{proc} split
  & Compiler-enforced purity with termination guarantees.
    No other listed language enforces purity at the declaration level.
    \\[4pt]
Caller-site ownership annotations
  & \lain{mov} and \lain{var} are visible at every call site, making
    ownership transfer explicit and auditable.
    \\[4pt]
Interval-based VRA
  & Polynomial-time static verification of array bounds and arithmetic
    constraints, without an SMT solver.
    \\[4pt]
\lain{in} operator for index constraints
  & \lain{idx in arr} establishes the bounds guard
    $0 \leq \mathtt{idx} < \mathtt{arr.len}$ in the containing scope.
    \\[4pt]
Bounded \lain{while} with \lain{decreasing}
  & Allows terminating loops in pure functions via programmer-supplied
    ranking functions, verified by the compiler.
    \\[4pt]
Non-consuming match (\lain{case \&expr})
  & Pattern matching that does not consume the scrutinee; the scrutinee
    is borrowed for the duration of the match body.
    \\[4pt]
\end{longtable}
